
---
title: 'Scrape contents of JOSS papers'
author: 'Mark Padgham'
date: '`r Sys.Date()`'
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: no
    theme: flatly
---

## Set up client for github graphql API

Also source a couple of helper functions
```{r gh-cli}
token <- Sys.getenv("GITHUB_GRAPHQL_TOKEN") # or whatever
gh_cli <- ghql::GraphqlClient$new (
    url = "https://api.github.com/graphql",
    headers = httr::add_headers (Authorization = paste0 ("Bearer ", token))
)
source ("get-joss-papers-extra-functions.R")
```


## Get all issue numbers of original JOSS submissions

```{r get_issue_labels, eval = FALSE}
library (ghrecipes)
issues <- get_issue_labels_state (owner = "openjournals", repo = "joss-reviews")
saveRDS (issues, file = "joss-reviews-all.Rds")
```

```{r get_issue_numbers, eval = FALSE}
library (ghrecipes)
library (dplyr)
issues <- readRDS ("joss-reviews-all.Rds") %>%
    filter (state == "CLOSED" &
            grepl ("accepted", .$labels) &
            repo == "joss-reviews") # They all are, but just in case
issue_nums <- issues$number
```

Those issues do not include the labels used to identify languages. Instead,
those can be scraped directly from the main JOSS page:
```{r}
library (magrittr)
library (rvest)
josslinks <- NULL
newlinks <- NA
pagenum <- 1
while (length (newlinks) > 0) {
    u <- paste0 ("https://joss.theoj.org/papers/in/R?page=", pagenum)
    pagenum <- pagenum + 1
    newlinks <- read_html (u) %>%
        html_nodes ("link") %>%
        html_attr ("href")
    newlinks <- newlinks [grep ("/papers/10.21105", newlinks)]
    josslinks <- c (josslinks, newlinks)
}

# Then transform those links to the actual software repo link from the
# corresponding JOSS page
get_repo <- function (link) {
    b <- read_html (link) %>%
        html_nodes ("a.btn") # link to repo is a "btn" class object
    i <- grep ("Software repository", html_text (b))
    html_attr (b [i], "href")
}
repolinks <- vapply (josslinks, function (i) get_repo (i), character (1),
                     USE.NAMES = FALSE)
links <- data.frame (joss = josslinks, repo = repolinks,
                     stringsAsFactors = FALSE)
saveRDS (links, file = "joss-repos-all.Rds")
```


## Get contents of `paper.md` from original repos


```{r get_papermd}
library (magrittr)
links <- readRDS ("./joss-repos-all.Rds")
links <- links [grepl ("github\\.com", links$repo), ]
txt <- paste0 ("Extracting paper.md entries from ", nrow (links),
               " repositories")
message (cli::rule (center = txt, line = 2, col = "green"))

# Get the `paper.md` file from the repo URL
get_papermd <- function (gh_cli, repo) {
    repo$repo <- gsub ("www\\.github\\.com", "github\\.com", repo$repo)
    url_cut <- strsplit (repo$repo, "/") [[1]]
    repo <- list (joss = repo$joss,
                  url = repo$repo,
                  raw = gsub ("github.com", "raw.githubusercontent.com",
                              repo [2]),
                  org = tail (url_cut, 2) [1],
                  repo = tail (url_cut, 1))

    files <- get_repo_files (gh_cli, repo$org, repo$repo)
    if (is.null (files))
        return (NA_character_)
    mdfile <- files [grep ("paper\\.md", files$name), ]
    if (nrow (mdfile) < 1)
        return (NA_character_)
    u <- paste0 (repo$raw, "/master/", mdfile$name [1])
    repo$md <- tryCatch (readLines (u, warn = FALSE),
                         error = function (e) NULL)
    return (repo)
}

#papers <- pbapply::pbapply (links, 1, function (i) get_papermd (gh_cli, i))
papers <- list ()
pb <- txtProgressBar (style = 3)
for (i in seq (nrow (links))) {
    papers [[length (papers) + 1]] <- get_papermd (gh_cli, links [i, ])
    setTxtProgressBar (pb, i / nrow (links))
}
close (pb)

papers <- papers [which (!is.na (papers))]
names (papers) <- vapply (papers, function (i) i$repo, character (1))
saveRDS (papers, file = "joss-papersmd-all.Rds")
```

Then format them as a single `.html` with index

```{r}
papers <- readRDS ("joss-papersmd-all.Rds")
out <- c ("---",
          "title: \"All R papers submitted to JOSS\"",
          "author: \"\"",
          "output:",
          "    html_document:",
          "        toc: true",
          "        toc_float: true",
          "        number_sections: false",
          "        theme: flatly",
          "---", "", "")

count <- 1
for (i in papers) {
    md <- i$md
    hdr <- grep ("^---$", md)
    if (length (hdr) > 1)
        md <- md [(hdr [2] + 1):length (md)]
    if (any (grepl ("^# References$", md)))
        md <- md [which (!grepl ("^# References$", md))]

    # Remove image links:
    md <- gsub ("\\.jpeg", "\\.jpg", md, ignore.case = TRUE)
    index <- grep ("\\.png|\\.jpg|\\.pdf", md, ignore.case = TRUE)
    for (j in index) {
        starts <- gregexpr ("\\!\\[", md [j]) [[1]]
        ends <- gregexpr ("\\.png\\)|\\.jpg)|\\.pdf\\)", md [j],
                          ignore.case = TRUE) [[1]] + 4
        for (k in seq_along (starts)) {
            if (starts [k] == 1 & ends [k] == nchar (md [j]))
                md [j] <- ""
            else if (starts [k] == 1 & ends [k] < nchar (md [j]))
                md [j] <- substring (md [j], ends [k] + 1, nchar (md [j]))
            else if (starts [k] > 1 & ends [k] == nchar (md [j]))
                md [j] <- substring (md [j], 1, starts [k] - 1)
            else
                md [j] <- paste0 (substring (md [j], 1, starts [k] - 1), " ",
                                  substring (md [j], ends [k] + 1, nchar (md [j])))
        }
    }

    ## change all secondary to tertiary, and primary to secondary headers
    md <- gsub ("^##\\s", "### ", md)
    md <- gsub ("^#\\s", "## ", md)

    # And remove any standard markdown header notations:
    md [grep ("^=+$", md)] <- ""

    out <- c (out, "", paste0 ("# ", count, ". ", i$repo), "",
              paste0 ("[", i$org, "/", i$repo, "](", i$url, ") [JOSS](",
                      i$joss, ")"), "",
              md)
    count <- count + 1
}
fname <- "joss-papersmd-all"
con <- file (paste0 (fname, ".md"), open = "w")
writeLines (out, con = con)
close (con)
```

Then render the markdown to html (takes a few seconds)
```{r}
rmarkdown::render(paste0 (fname, ".md"), output_file = paste0 (fname, ".html"))
```




